DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-09-11 2015CompleteDerivation/completeDerivation.tex
First part for quantum

\section{Composite systems and reducibility}

@@

\subsection{Homogeneous decomposable systems and vector spaces}
The notion that a system is decomposable means the states are equipped with a rule of composition that allows one to write $\mathcal{c}=\mathcal{c}_1+\mathcal{c}_2$: the composite system is the sum of its parts. For example, the state of a ball is equal to the state of its top and bottom parts.

@@

\begin{prop}\label{reducible_state_space}
The state space $\mathcal{C}$ for a decomposable homogeneous material is an additive abelian (i.e. commutative) group.
\end{prop}

@@

\begin{justification}
We claim $\mathcal{C}$ is an abelian group. $\mathcal{C}$ is the state space for a decomposable homogeneous system. $\mathcal{C}$ is an abelian group  by \ref{reducible_state_space}.

@@

\section{Quantum systems}

We now focus on a different type of homogeneous material: one that is irreducible. That is, the state and the evolution of the whole does not tell us anything about the state and the evolution of the parts. As the internal dynamics is undetermined, we'll find that each irreducible element is suitably described by the standard framework of quantum mechanics.

TODO

>>>>

\section{Composite systems and reducibility}
\label{sec:reducibility}

@@

\subsection{Homogeneous decomposable systems and real vector spaces}
The notion that a system is decomposable means the states are equipped with a rule of composition that allows one to write $\mathcal{c}=\mathcal{c}_1+\mathcal{c}_2$: the composite system is the sum of its parts. For example, the state of a ball is equal to the state of its top and bottom parts.

@@

\begin{prop}\label{prop:abelian_group}
The state space $\mathcal{C}$ for a decomposable homogeneous material is an additive abelian (i.e. commutative) group.
\end{prop}

@@

\begin{justification}
We claim $\mathcal{C}$ is an abelian group. $\mathcal{C}$ is the state space for a decomposable homogeneous system. $\mathcal{C}$ is an abelian group  by \ref{prop:abelian_group}.

@@

\section{Irreducible systems}

In this section we turn to a different case in which the relationship between the whole material and its parts is different from the one described in section \ref{sec:reducibility}. The system is still infinitesimally decomposable, made of infinitesimal parts as before, but this time it is irreducible. That is: the state and the evolution of the whole does not tell us anything about the state and the evolution of the parts. As the internal dynamics is undetermined, we'll find that each irreducible element is suitably described by the standard framework of quantum mechanics.

\subsection{Quantized material}

In section \ref{sec:reducibility} we introduced the idea of a classical material, one that is homogeneous and infinitesimally reducible, which represented a way to relate the state of the whole system to the state of its parts. In this section we explore an alternative case.

We will call a \emph{quantized material} a homogeneous material that is finitely reducible. That is, as we decompose the system we can only assign a state to finite amounts of material, which we call \emph{quantum particle}. As the material is homogeneous, all its particles carry the same amount of material and are described by the same state space. Each quantum particle is irreducible: its state does not tell us anything about the state of its parts. While this idea should be familiar enough, it is best to iron out some details to avoid possible misconceptions.

The first possible source of confusion is between divisibility, the idea that an object can be divided into parts, and reducibility, the idea that the state of the whole is equivalent to the state of the parts. Most of the time these two concepts overlap, but they are separate as we can see in the following examples.

Consider the a planarian flatworm. If we cut it into multiple pieces, each part will grow into a new worm. Yet, the state of each worm is not the state of multiple worms put together. The planarian is divisible into multiple worms but it is not reducible to multiple worms. Consider a magnet. Its state is described, to a first approximation, by the position of the north and south poles. Yet, we cannot divide the magnet into a north and south pole. The magnet is reducible to the poles but is not divisible into the poles.

Consider a muon. It will decay into an electron and two neutrinos. But the state of a muon is not the state of an electron and two neutrinos. The muon is divisible into the three particles but it is not reducible into them. Consider a proton. Its state is given by the state of its three valence quarks. Yet, we cannot divide the proton into the three quarks. The proton is reducible to quarks, but divisible into quarks.

The fundamental property of a quantum particle, as introduced here, is not indivisibility: it's irreducibility. It is the state that cannot be further partitioned, not the material itself.

Even more subtle is the difference between divisibility and decomposability, the idea that the object has parts. While we are agnostic on whether a quantum particle can be divided, we are implicitly assuming that it can be decomposed. Irreducibility describes the relationship between the state of the whole and the state of the parts, therefore parts are implicit in the definition. While these parts must exist, we cannot isolate them and act on them independently: if we were able to do that we could assign them a state.

While we must be able to talk about parts of a quantum particles, not all questions will make sense. Can we ask what fraction of a quantum particle is in a particular region of space? Yes, since the particle is decomposed some part may be inside the volume. Can we ask where that same fraction will be at a later time? No, since the particle is irreducible we cannot describe the evolution of a part.

The proper way to think of a quantized material is therefore the following: a continuum of infinitesimal parts clumped together to form quantum particles, the smallest amount of material to which we can assign a state. With this picture in mind, we can see, even before developing the math, how some features typical of quantum behaviors emerge.

The first feature is the importance of statistics in the quantum particle state. Consider a part of a quantum particle in an initial state. As we cannot track its motion, it can be any part of the particle in its final state. That is, the motion of a part of a particle is a random walk within the evolution of a particle.

The second feature is non locality. Suppose we have a quantum particle state distributed in space. Suppose we exert a force only on a subregion of its distribution. As the particle is irreducible, we cannot interact with only part of it: we affect also the part where no force is exerted. We cannot simply interact with the part at a specific region, as this would allow to assign a state to that part.

Wave particle duality should be clear at this point. The quantum particle behaves like a wave in the sense that its parts form a distribution in space. Yet, it behaves as a single unit, as the state is one unit and cannot be separated.

The most striking feature, though, is that we can predict the functional form for particle decay. Suppose a quantum particle spontaneously decays into two or more particles. Such process cannot be deterministic as the d.o.f. increase, therefore is not determined by the state of the quantum particle. As the decay is spontaneous, it is not triggered by the environment (i.e. external state). It is therefore triggered by the internal unstated dynamic of the parts. As we assumed the internal dynamic is unknown, we cannot learn anything about it by looking at the product of the decay. In particular, all states have to have equal chance of decay or we could tell that the internal dynamic is more stable for some states. The probability of decay is constant in time. The probability of survival follows the exponential distribution.

As we see qualitatively how the features typical of quantum mechanics naturally emerge from this picture, we are ready to make this more quantitatively and mathematically precise.

\subsection{Quantum particle and complex vector spaces}

Turning these concept into a precise mathematical formulation is tricky as we need to formally characterize the unstated part which is not physically distinguishable. While many previous results do not apply, the classical derivation still can serve as a template.

As in the classical case, the quantum particle is thought to be decomposable: different parts will be in different places, the evolution of the whole is the composition of the evolution of parts. Therefore \ref{prop:abelian_group} applies and the state space $\Psi$ of an irreducible material is an abelian group.

Note, though, that we can't think of the parts as independent components. As the unstated part is statistical in nature, it will combine as statistical variables do. Suppose $X$ and $Y$ are two random variables, $\sigma^2_{X+Y} = \sigma^2_{X} + \sigma^2_{Y} + 2 \, \sigma_{X} \sigma_{Y} \rho_{X,Y}$, where $\sigma^2$ is the variance, $\sigma$ is the standard deviation, and $\rho_{X,Y}=COV(X,Y)/\sigma_{X} \sigma_{Y}$ is the Pearson correlation coefficient. As $-1\leq\rho_{X,Y}\leq+1$, the total variance can be greater or smaller then the sum of the independent parts, and the strength of the uncertainty within the unstated part is something physically distinguishable.

Therefore, while the motion of the unstated part is not itself physically relevant, the correlation between the motion of the unstated part of two combined parts is very relevant. We'll see how interference can be understood in this context: it is constructive if the randomness is correlated while it is destructive if it is anti-correlated.

So, how can we characterized these relationship between different configurations of the unstated parts? We can look at the classical derivation for guidance. In that case, we looked at the set of transformation that increase/decrease the amount of material. Those are the only type of transformation that we can apply on a classical system that do not change the overall state given by the distribution.

As the classical material is homogeneous, permutations between classical particles in the same state $\mathcal{s}$ do not matter: the ordering is unstated. In a way, increasing/decreasing the amount of material increases/decreases the unstated part of the system.

For the irreducible material we follow the same strategy: we find the state transformations that affect the unstated part while keeping the overall state, the distribution, unchanged.

Suppose we have $\psi_1$ a state for an irreducible material characterized by one d.o.f. The material will be spread over the configurations of the infinitesimal fragments. As the material is homogeneous, the material is spread evenly among the such possible configuration. Counting configurations is therefore equivalent to counting amount of material. Each fragment would be identified by a pair of independent random variables $(A, B)$, as a d.o.f. is two-dimensional. The shape of the distribution and the actual values are not physically distinguishable: these are just dummy variables we assign to count the fragment configurations. We choose A and B to be uniform distributions from 0 to 1, so that the area is a square of unit size, which represent the unit of unstated  configurations the material is spread over.

Suppose we have another state $\psi_2$, where only unstated part is changed. The unstated part can also be characterized by a square formed by two independent uniform distributions (C, D). Counting states on (C,D) is equivalent to counting them on (A,B), as they are simply different arrangements. As all distributions are uniform, they can only be connected by a linear operation. That is: $C=aA+bB$ and $D= -bA + aB$. We can write $\psi_2 = \tau(a,b) \psi_1$ as $a$ and $b$ are the only parameters of the transformation.

The transformation $\tau(a,b)$ increases/decreases the area by the factor $a^2+b^2$ and rotates the square by $arctan(x/y)$. If we set $c=a+\imath b$ those same quantities can be expressed as $\rho^2=|c|^2$ and $\theta = arg(z)$, the modulus and the phase of a complex number. In other words: complex numbers provide a representation for the group of rotations and magnitude rescaling.

For multiple independent d.o.f., we have one square each. As the material is irreducible, all areas need to be always the same since relative changes would allow to learn something about the motion of the infinitesimal fragments. This means that the transformation $\tau(c)$ has to be the same for all d.o.f. In other words, the group of transformations is the same regardless of the number of independent d.o.f.

Mathematically, the state space $\Psi$ for an irreducible material is a complex vector space.

\begin{prop}\label{prop:complex_vector_space}
	The state space $\Psi$ for an irreducible material is a vector space over $\mathbbm{C}$.
\end{prop}

\begin{justification}
	TODO
\end{justification}

If $\mathcal{Q}$ is the space that defines our units, a state $\psi$ must tell us for each point $\mathcal{q} \in \mathcal{Q}$ how much material is there and the the correlation of the unstated part. That is: the state will be identified by $\psi : \mathcal{Q} \rightarrow \mathbbm{C}$, a complex function over the $\mathcal{Q}$.

The square modulus at each point will give us the density of the material. As in the classical case, such density mush be integrable to give us the total amount of material. Therefore the function $\psi \in L^2(\mathcal{Q}, \mu)$ is square integrable. The norm $|\psi|^2 = \int_{\mathcal{Q}} |\psi(q)|^2 dq$ will give us the amount of material associated with the state.

As the material is homogeneous, all particles of the material are made of the same amount. We can set, by convention, the amount of material for a particle to be unitary, therefore quantum particle states are the subset of $\Psi$ such that $|\psi|^2 = 1$. Note that changing the phase by the same amount everywhere is only changes the unstated part. Such change does not change the state, therefore distribution whose differ by a total phase are not physically distinguishable and represent the same physical state.

\begin{prop}\label{prop:wavefuntion}
	The state space $\Psi$ for an irreducible material is isomorphic to the space of Lebesgue square integrable differentiable functions over $\mathcal{Q}$. That is $\Psi \cong C^1(\mathcal{Q}) \cap L^1(\mathcal{Q}, \mu)$. The state space for a quantum particle is the projective space $\mathsf{P}(\Psi)$.
\end{prop}
\begin{justification}
	TODO
\end{justification}

The observation made after \ref{prop:differentiable_manifold} in classical case are still valid: $\Psi$ is not a complete metric space.

\subsection{Irreducibility}

\subsection{Quantum mechanics}
